%% Latex lort {{{1
\documentclass[a4paper,11pt,fleqn]{article}

\usepackage{a4wide}
\usepackage[danish]{babel}
\usepackage{t1enc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{times}
\usepackage{theorem}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{lscape}
\usepackage{clrscode}
\title{Data Structures Exam Notes}
\author{Asser Schrøder Femø}

\begin{document}
\maketitle
%% }}}1

\section{Selection}

\begin{itemize}
\item Order statistic: $k$th order statistic = $k$th largest element out of $n$
elements

\item Selection problem: Report the $k$th largest element (ie. $k$th order
statistic)

\item Trivial solution: Sort ($\Theta(n \lg n)$) then select

\item Randomized select: Quicksort but only recurse on half the set. Expected
(due to randomization) running time $O(n)$, worst case $\Theta(n^2)$

\item Selection by sieving: 
\begin{enumerate}
\item Divide into groups of 5 elements - $O(n)$
\item Find median on each group by insertion sort - $O(n)$
\item Recursively find median of each group - $T(3n/10)$
\item Partition around median-of-medians - $O(n)$
\item Recursively find the $i$th smallest element on the relevant side - $T(7n/10 + 6)$
\end{enumerate}

Recurrence: 

\begin{displaymath}
T(n) = \left\{ \begin{array}{ll}
 O(1) & \textrm{if $n<140$}\\
  T(\lceil n/5 \rceil) + T(7n/10 + 6) + O(n) & \textrm{if $n \geq 140$}
     \end{array} \right.
\end{displaymath}

Solution: $T(n) = O(n)$

\end{itemize}

\section{Amortized analysis}

\begin{enumerate}
\item Aggregate analysis: When a sequence of $n$ operations take $T(n)$ time then the
amortized cost per operation is $T(n)/n$

\item Accounting method: Assign \emph{charges} to different operations. Each
operation ``charges'' or credits the data structure.

\item Potential method: Pick a function $\Phi(D)$ that maps the configuration of
$D$ to a a real number, called the potential of $D$.

	\begin{enumerate}
	\item $\Phi_i \geq \Phi_0$ for all $i \in {1,\ldots,n}$.
	\item $t_i \leq k (c_i - \Delta_i)$ where $\Delta_i = \Phi_i - \Phi_{i-1}$, $k$
	constant
	\end{enumerate}
\end{enumerate}

\section{Number systems}

\begin{itemize}
\item Incrementing binary numbers
\item Skew binary numbers (decrement in $O(1)$ time due to redundant 2)
\item Analytical ties to other data structures
\end{itemize}

\section{Fibonacci heaps}

Amortized running times:

\begin{description}
\item[Insert] $\Theta(1)$ (also actual cost): Insert as root node.

\item[Find-min] $\Theta(1)$: Return minimum element from pointer.

\item[Extract-Min] $O(\log n)$: Add children to root list, then consolidate into
binomial trees.

\item[Union] $\Theta(1)$: Append root nodes.

\item[Decrease-Key] $\Theta(1)$: Mark parent, cut node and make it a root. If
parent just lost its second child cut it and make it a root.

\item[Delete] $O(\log n)$: Run decrease-key then extract-min.
\end{description}

Potential function: $\Phi(H) = t(H) + 2m(H)$

Maximum degree $D(n) = O(\log n)$. Proof by induction using the Fibonacci
sequence.

\section{van Emde Boas Trees}

\begin{itemize}
\item Recursive Structure
\item Numbers in universe $u$
\item $O(\log \log u)$ running time for all operations: Minimum, Maximum, Successor,
Predecessor, Insert, Delete, Lookup.
\item Space requirement: $P(u) = O(u)$
\end{itemize}

\section{Hashtables}

Support operations: Insert, Search, Delete.

\begin{description}
\item[Hash function] $h$ maps elements to slot numbers: $h : U \rightarrow {0,1,\ldots,m-1}$
\item[Collisions] resolved by chaining. Avg case time $\Theta(1 + \alpha$ where
$\alpha$ is the table load factor.
\item[Universal Hashing] randomly chosen hash function independently of keys to
avoid worst-case behavior. Expected $\Theta(n)$ time for any sequence of $n$
operations.
\item[Perfect hashing] 
\end{description}

\section{External-memory data structures}

B-trees.

\begin{itemize}
\item Properties: $n$ keys in each node $x$ stored in ascending order. Each
internal node has $n+1$ pointers to its children. Keys separate ranges stored in
each subtree. All leaves have same depth = height of the B-tree $h$. 
\item Minimum degree $t \geq 2$ - every node must have at least $t - 1$ keys and
at least $t$ children, and at most $2t - 1$ keys and $2t$ children.
\item Height of B-tree: $h \leq \log_t \frac{n+1}{2}$
\item Searching: Analogue to binary search tree, disk accesses = $O(h) =O(\log_t n)$
\item Insert: Traverse the tree, split full nodes as necessary and insert in
nonempty leaf node with $O(h)$ disk accesses.
\item Deletion: $O(h)$
\end{itemize}

\section{Self-adjusting data structures}

\begin{itemize}
\item Search trees (splay trees)
\item Also: Pairing heaps :-)
\item Potential function
\begin{enumerate}
\item $s(x)$ size of a node = the number of nodes in its subtree including itself
\item $r(x)$ rank of a node = $\log s(x)$
\item $\Phi(D) = \sum^n_{i=0} r(n_i)$
\end{enumerate}
\end{itemize}

\end{document}
